---
title: "Extraction of March 2020 US census data on population ocupation and income "
author: "Group A - Autumn Class:  Junylou Daniel, Oana Damian, Robin Mathew, Torsten Meyer"
date: '01 November 2020'
url: https://archive.ics.uci.edu/ml/datasets/Adult
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	message = FALSE,
	warning = FALSE
)
```



### About 

Extraction of US census data from the Current Population Survey (CPS), Annual Social and Economic (ASEC) Supplement as at March 2020

The data is made available by the US Census Bureau <https://www.census.gov/data/datasets/time-series/demo/cps/cps-asec.2020.html>

Data description: <file:///C:/Users_Folders/YORK_MLcertificate/Assignment2/asecpub20csv/ASEC2020ddl_pub_full_fields_description.pdf>

Data dource: <https://www2.census.gov/programs-surveys/cps/datasets/2020/march/asecpub20csv.zip>

To run the extraction code download the following 3 files on your local machine:

files 1 & 2 unzip csv files "hhpub20.csv" and "pphub20.csv"  from  <https://www2.census.gov/programs-surveys/cps/datasets/2020/march/asecpub20csv.zip>

file 3 "INPUT_variables_map.csv" from  <https://github.com/oanada/Assignment2>

The result of the data extraction is uploaded and available for download as "US_Census_2020.csv" in the public repository <https://github.com/oanada/Assignment2>


```{r, message = FALSE}
## pls. download and unzip csv files "hhpub20.csv" and "pphub20.csv"  from  https://www2.census.gov/programs-surveys/cps/datasets/2020/march/asecpub20csv.zip
## pls download "INPUT_variables_map.csv" from  https://github.com/oanada/Assignment2


# Pls input below the local folder where the three files ("hhpub20.csv", "pphub20.csv", "INPUT_variables_map.csv") have been downloaded
# The processed data extraction will be saved in the same folder

INPUT_local_folder="C:/Users_Folders/YORK_MLcertificate/Assignment2/asecpub20csv/"


### read variables mapping
### ===================================================
INPUT_varmap=read.csv(paste0(INPUT_local_folder,"INPUT_variables_map.csv"), header = TRUE)


### Load persons data 
### ===================================================

INPUT_pp_retained_fields=      c('PERIDNUM','PH_SEQ', 'A_FNLWGT', 'A_AGE',    'A_CLSWKR',  'A_HGA',     'A_MARITL',       'A_PFREL',       'PRDTRACE', 'PENATVTY',       'A_MJOCC',      'A_USLHRS',       'WSAL_VAL'   ,   'CAP_VAL' ,'A_SEX','A_HSCOL')
INPUT_pp_retained_fields_names=c('PERIDNUM','H_SEQ',  'fnlwgt',   'age',      'workclass', 'education', 'marital_status', 'relationship',  'race',     'native_country', 'occupation',   'hours_per_week', 'Wages_n_salary', 'capital_gains','sex','School_enrolment')
INPUT_pp_fields_labeled=c('PENATVTY','PRDTRACE','A_PFREL','A_MARITL','A_HGA','A_CLSWKR','A_MJOCC','A_SEX','A_HSCOL')


INPUT_pp=read.csv(paste0(INPUT_local_folder,"pppub20.csv"), header = TRUE)
# WSAL_VAL - total wage and salary earnings (combined amounts in ern-val, if  ern-srce=1, and ws-val)


#A_USLHRS= How many hrs per week does ... usually work at this job? Universe: All Persons
# Values: -4 = Hours vary
# -1 = Not in universe
# 00 = None, no hours
# 01-99 = Entry



# retain only a subset of available fields
INPUT_pp=INPUT_pp[,INPUT_pp_retained_fields]


# remove children
INPUT_pp=INPUT_pp[!(INPUT_pp$A_HGA==0 & INPUT_pp$A_AGE<=16),]


# remove ppl who have no occupation or are children A_MJOCC 
INPUT_pp=INPUT_pp[!(INPUT_pp$A_MJOCC==0),]

# remove ppl who have A_FNLWGT=0 
INPUT_pp=INPUT_pp[!(INPUT_pp$A_FNLWGT==0),]


# replace variable values with the labels
for(idx_var in 1:length(INPUT_pp_fields_labeled))
{ TEMP_varmap=INPUT_varmap[(INPUT_varmap[,1]==INPUT_pp_fields_labeled[idx_var]),c(2,4)]
  colnames(TEMP_varmap)=c(INPUT_pp_fields_labeled[idx_var],'Label')
  INPUT_pp<- merge(x = INPUT_pp, y = TEMP_varmap, by = INPUT_pp_fields_labeled[idx_var], all.x = TRUE)
  INPUT_pp[,c(INPUT_pp_fields_labeled[idx_var])]=INPUT_pp$Label
  INPUT_pp = INPUT_pp[,!(names(INPUT_pp) %in% c("Label"))]
  
}


# rename headers
for(idx_var in 1:length(INPUT_pp_retained_fields))
{colnames(INPUT_pp)[colnames(INPUT_pp) == INPUT_pp_retained_fields[idx_var]] <- INPUT_pp_retained_fields_names[idx_var]
}




### Load households data
### ===================================================

INPUT_hh_retained_fields=      c('H_SEQ', 'GEDIV',  'GESTFIPS','GTCBSASZ')
INPUT_hh_retained_fields_names=c('H_SEQ', 'region', 'state',   'metropoly_size')
INPUT_hh_fields_labeled=      c( 'GEDIV',  'GESTFIPS','GTCBSASZ')

INPUT_hh=read.csv(paste0(INPUT_local_folder,"hhpub20.csv"), header = TRUE)

INPUT_hh=INPUT_hh[,INPUT_hh_retained_fields]

# replace variable values with the labels
for(idx_var in 1:length(INPUT_hh_fields_labeled))
{ TEMP_varmap=INPUT_varmap[(INPUT_varmap[,1]==INPUT_hh_fields_labeled[idx_var]),c(2,4)]
  colnames(TEMP_varmap)=c(INPUT_hh_fields_labeled[idx_var],'Label')
  INPUT_hh=merge(x = INPUT_hh, y = TEMP_varmap, by = INPUT_hh_fields_labeled[idx_var], all.x = TRUE)
  INPUT_hh[,c(INPUT_hh_fields_labeled[idx_var])]=INPUT_hh$Label
  INPUT_hh = INPUT_hh[,!(names(INPUT_hh) %in% c("Label"))]
  
}


# rename headers
for(idx_var in 1:length(INPUT_hh_retained_fields))
{colnames(INPUT_hh)[colnames(INPUT_hh) == INPUT_hh_retained_fields[idx_var]] <- INPUT_hh_retained_fields_names[idx_var]
}




### join household geographical data with people data
### ===================================================
INPUT_pp_hh<- merge(x = INPUT_pp, y = INPUT_hh, by = "H_SEQ", all.x = TRUE)
write.csv(INPUT_pp_hh, paste0(INPUT_local_folder,"US_Census_2020.csv"))

```


